Accessing biological data in R with semantic web technologies

### 2015-02-18 article.tex
Updated for today's rrdf and rrdflibs versions

\newpage
\begin{abstract}
\noindent
\textbf{Background}. Semantic Web technologies are increasingly used in biological database systems.
The improved expressiveness shows advantages in tracking provenance and allowing knowledge to be
more explicitly annotated. The list of semantic web standards needs a complementary set of tools to
handle data in those formats to use them in bioinformatics workflows. \\
\textbf{Methods}. The approach proposed in this paper uses the Apache Jena library to create an
environment where semantic web technologies can be used in the statistical environment R. The
code is exposed as two R packages available from the Comprehensive R Archive Network (CRAN). 
The RJava library and a custom convenience class is used to bridge between R and the Jena library.\\
\textbf{Results}. We here present three examples showing how the Resource Description Framework
(RDF) and SPARQL query standards can be employed in R. The first example takes input on BRCA1 SNPs
from a BioMart and converts this into a RDF data set. The second example runs a query on an
experimental remote SPARQL end point provided by Uniprot, and searches textual annotations of
proteins encoded by the BRCA1 gene. The third example shows how the package can be used to handle
RDF returned by OpenTox web services. \\
\textbf{Discussion}. The two provided library bring basic semantic web technologies to R.
This paper only shows examples from the biology domain, but we believe the approaches are generally
applicable. It currently only exposes a subset of key Apache Jena, but the rrdf package
makes it easy to make more of the library's functionality, such as the shortest path finding.
The rrdf libraries are freely available from the CRAN under the 
Affero GNU Public License version 3: \url{http://cran.r-project.org/web/packages/rrdf/}.
\end{abstract}

@@

The rrdf packages provide RDF and SPARQL functionality by exposing that functionality from the
Apache Jena library~\cite{Jena}. The rrdflibs package contains the Apache Jena libraries and
nothing more. This package is close to 7~MiB large but is expected to change only with major new
Jena releases. The rrdf package, however, contains the R functions that wrap around the Jena
functionality and convert data structures where needed. This package is about 112~kiB and can
now be updated without users having to download
the same Jena libraries repetitively. Using this approach,
the Comprehensive R Archive Network (CRAN) server has the least amount of download throughput.
The Jena functionality is accessible using the rJava library~\cite{RJava} which handles loading
of the Java archives and provides methods to instantiate classes and call methods.

The source code of the package is available from GitHub at \url{http://github.com/egonw/rrdf/} and
binary packages are available from CRAN at \url{http://cran.r-project.org/web/packages/rrdf/}.
The first patch is from March 2011 while the most recent patches are from late 2013 when the
rrdflibs package was updated for Apache Jena 2.11.

>>>>

\newpage
\begin{abstract}
\noindent
\textbf{Background}. Semantic Web technologies are increasingly used in biological database systems.
The improved expressiveness shows advantages in tracking provenance and allowing knowledge to be
more explicitly annotated. The list of semantic web standards needs a complementary set of tools to
handle data in those formats to use them in bioinformatics workflows. \\
\textbf{Methods}. The approach proposed in this paper uses the Apache Jena library to create an
environment where semantic web technologies can be used in the statistical environment R. The
code is exposed as two R packages available from the Comprehensive R Archive Network (CRAN). 
The RJava library and a custom convenience class is used to bridge between R and the Jena library.\\
\textbf{Results}. We here present three examples showing how the Resource Description Framework
(RDF) and SPARQL query standards can be employed in R. The first example takes input on BRCA1 SNPs
from a BioMart and converts this into a RDF data set. The second example runs a query on an
experimental remote SPARQL end point provided by Uniprot, and searches textual annotations of
proteins encoded by the BRCA1 gene. The third example shows how the package can be used to handle
RDF returned by OpenTox web services. \\
\textbf{Discussion}. The two provided library bring basic semantic web technologies to R.
This paper only shows examples from the biology domain, but we believe the approaches are generally
applicable. It currently only exposes a subset of key Apache Jena, but the rrdf package
makes it easy to make more of the library's functionality, such as the shortest path finding.
The rrdf packages are freely available from GitHub under the 
Affero GNU Public License version 3: \url{https://github.com/egonw/rrdf/}.
\end{abstract}

@@

The rrdf packages provide RDF and SPARQL functionality by exposing that functionality from the
Apache Jena library~\cite{Jena}. The rrdflibs package contains the Apache Jena libraries and
nothing more. This package is about 7~MiB large but is expected to change only with major new
Jena releases. The rrdf package, however, contains the R functions that wrap around the Jena
functionality and convert data structures where needed. This package is about 112~kiB and can
now be updated without users having to download
the same Jena libraries repetitively. Using this approach, the download size is minimized.
The Jena functionality is accessible using the rJava library~\cite{RJava} which handles loading
of the Java archives and provides methods to instantiate classes and call methods.

The source code of the package is available from GitHub at \url{http://github.com/egonw/rrdf/}.
Binary packages are available here too and can be installed easily from GitHub using the 
devtools package~\cite{devtools}. The first patch of the library is from March 2011 while the
most recent patches are from late 2013 when the rrdflibs package was updated for Apache Jena 2.11.
Early 2015 the rrdflibs package was updated to Apache Jena 2.12.1.
